Progress Report on NHANES Data

### 2023-10-03 Docs/ProgressReportAppl.Rmd
fix laplace failures

---
title: "Progress Report on NHANES Data"
author: "Ying Jin"
date: "2023-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# NHANES data

## Data overview

@@

## Out-of-sample prediction

@@

Now I am going to use one subject as an example to go over the approximation procedure:

@@

```{r, class.source = "fold-show"}
## model
Model <- function(parm, Data){
  xi <- parm[Data$pos.xi]

@@

# Full data application output

## fGFPCA

- Still need to project eigenfunctions to the full grid

@@

```{r message=FALSE, warning=FALSE}
df_test_full <- df %>% 
  filter(id %in% test_id) %>%
  filter(!id %in% skip_id) %>%
  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin"))

@@

### Numeric problems

- Three subjects had failed approximation, all happened with the shortest observed track
- Does it have to do with still the length of observed track? Or the arbitrary choices in the binning procedure (bin width). 
- Per last meeting, adding some details on laplace approximation

```{r, fig.height=3, fig.width=9}
df %>% filter(id %in% skip_id) %>% 
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.2)+
  facet_wrap(~id)+
  geom_vline(xintercept = c(540))
```

@@

```{r}
# up to 540
J <- 1440
auc_t540_adglmm <- adglmm_pred_t540$newdata2 %>%
  filter(!id %in% skip_id) %>%
  mutate(sind=sind*J) %>%
  mutate(window = cut(sind, breaks = c(540, 780, 1020, 1260, 1440), 
                      labels = c("9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
auc_t540_adglmm <- split(auc_t540_adglmm, f=auc_t540_adglmm$window)
auc_t540_adglmm <- auc_t540_adglmm[1:3] %>% 
  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
auc_t540_adglmm <- lapply(auc_t540_adglmm, function(x){x@y.values[[1]]}) %>% unlist()

# up to 780
auc_t780_adglmm <- adglmm_pred_t780$newdata2 %>%
  filter(!id %in% skip_id) %>%
  mutate(sind=sind*J) %>%
  mutate(window = cut(sind, breaks = c(780, 1020, 1260, 1440), 
                      labels = c("1pm-5pm", "5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
auc_t780_adglmm <- split(auc_t780_adglmm, f=auc_t780_adglmm$window)
auc_t780_adglmm <- auc_t780_adglmm[1:2] %>% 
  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
auc_t780_adglmm <- lapply(auc_t780_adglmm, function(x){x@y.values[[1]]}) %>% unlist()

# up to 1020
auc_t1020_adglmm <- adglmm_pred_t1020$newdata2 %>%
  filter(!id %in% skip_id) %>%
  mutate(sind=sind*J) %>%
  mutate(window = cut(sind, breaks = c(1020, 1260, 1440), 
                      labels = c("5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
auc_t1020_adglmm <- split(auc_t1020_adglmm, f=auc_t1020_adglmm$window)
auc_t1020_adglmm <- auc_t1020_adglmm[1] %>% 
  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
auc_t1020_adglmm <- lapply(auc_t1020_adglmm, function(x){x@y.values[[1]]}) %>% unlist()

@@

## Compare fGFPCA and GLMMadaptive

@@

@@

- What would be the reference method to compare our performance to? Original GLMMadaptvie perhaps is feasible on the entire dataset. But what should we use if we wanna do sub-sample comparison? Should we even do that for this data application (or simulation alone)? 
- There are still three participants with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects. 
- Is interpolation really necessary? It does not improve performance very much but consumes time. Also, should I try more sophisticated methods for grid extension? 
- Should I calculate subject-wise AUC and average them? But some subjects are all zero within the prediction window, making AUC for this single subjects impossible to calculate...

>>>>

---
title: "Progress Report on NHANES Data"
author: "Ying Jin"
date: "2023-10-03"
output: 
  html_document:
    self_contained: no
    number_sections: true
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    font: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(516)

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
library(mgcv)
library(splines)
theme_set(theme_minimal())
```

```{r}
# load data
df <- read_rds(here("Data/nhanes_bi.rds"))

# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
source(here("Code/OutSampMLE.R"))
# source(here("Code/OutsampBayes.R"))
```

# Data overview

@@

# Fit fGFPCA model

## Bin data

- Bin every 10 observations

```{r, message=FALSE}
# bin data
bin_w <- 10 # bin width
n_bin <- J/bin_w # number of bins
brks <- seq(0, J, by = bin_w) # cutoff points
mid <- (brks+bin_w/2)[1:n_bin] # mid points

df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
df$bin <- as.numeric(as.character(df$bin))
# unique(df$bin)

df %>% 
  filter(SEQN %in% rand_id) %>%
  group_by(SEQN, bin) %>%
  summarise(num = sum(Z)) %>%
  ggplot()+
  geom_point(aes(x=bin, y=num), size = 0.5)+
  facet_wrap(~SEQN)+
  labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")
```
## Data split

- 80% (7010) subjects for training, 20% (1753) for out-of-sample prediction

```{r}
# head(df)
df <- df %>% rename(id=SEQN, Y=Z)

train_id <- sample(unique(df$id), size = N*0.8)
test_id <- setdiff(unique(df$id), train_id)

train_df <- df %>% filter(id %in% train_id)
test_df <- df %>% filter(id %in% test_id)
```

## Local GLMM

- Used glmer with PIRLS (nAGQ=0) for local GLMMs step
- Fit on the training set

```{r, cache=TRUE}
# fit model on the training set
train_bin_lst <- split(train_df, f = train_df$bin)

# local GLMM and estimate latent function
# use PIRLS (nAGQ=0) to avoid near-unidentifiability issues 
t1=Sys.time()
df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
t2= Sys.time()
t_local_glmm <- t2-t1 
```

```{r}
df_est_latent <- bind_rows(df_est_latent) 
# head(df_est_latent)

# example estimated latent function
rand_id <- sample(train_id, 4)

df_est_latent %>% 
  filter(id %in% rand_id) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id, col = "estimated"))+
  geom_point(aes(x=sind, y = Y, group = id, col = "outcome"), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

## FPCA

```{r}
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
# head(uni_eta_hat)

mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                         nrow=length(train_id), 
                         ncol=n_bin, byrow = F) 
# row index subject, column binned time
# dim(mat_est_unique)

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
t2 <- Sys.time()
t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model

# dim(fpca_mod$efunctions) # 27 eigenfunctions
K <- 4
```

```{r}
plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
```

```{r, fig.height=8, fig.width=8}
par(mfrow=c(2,2))

plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
```

```{r}
# plot correlation matrix
heatmap(cov2cor(fpca_mod$VarMats[[1]]), Rowv = NA, Colv = NA, main = "Correlation")
```

## Debias (step 4)

According to the fGFPCA paper, in this step we do two things:

1. Project the eigenfunctions with B-spline basis back to the original grid
2. Debias the eigenvalues with GLMM (mgcv::bam, mgcm::gamm or gamm4:gamm) to be used for Laplace approximation

Becuase I would like to focus this report on subjects with failed Laplace approximation, I will not do these two things for now.

# Out-of-sample prediction

@@

@@

## One subject example

Now I am going to use one subject as an example to go over the approximation procedure:
```{r, class.source = "fold-show"}
## model
Model <- function(parm, Data){
  xi <- parm[Data$pos.xi]

@@

## Full data application output

### fGFPCA

@@

```{r message=FALSE, warning=FALSE}
df_test_full <- df %>% 
  filter(id %in% test_id) %>%
  # filter(!id %in% skip_id) %>%
  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin"))

@@

@@

```{r}
# up to 540
J <- 1440
auc_t540_adglmm <- adglmm_pred_t540$newdata2 %>%
  # filter(!id %in% skip_id) %>%
  mutate(sind=sind*J) %>%
  mutate(window = cut(sind, breaks = c(540, 780, 1020, 1260, 1440), 
                      labels = c("9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
auc_t540_adglmm <- split(auc_t540_adglmm, f=auc_t540_adglmm$window)
auc_t540_adglmm <- auc_t540_adglmm[1:3] %>% 
  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
auc_t540_adglmm <- lapply(auc_t540_adglmm, function(x){x@y.values[[1]]}) %>% unlist()

# up to 780
auc_t780_adglmm <- adglmm_pred_t780$newdata2 %>%
  # filter(!id %in% skip_id) %>%
  mutate(sind=sind*J) %>%
  mutate(window = cut(sind, breaks = c(780, 1020, 1260, 1440), 
                      labels = c("1pm-5pm", "5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
auc_t780_adglmm <- split(auc_t780_adglmm, f=auc_t780_adglmm$window)
auc_t780_adglmm <- auc_t780_adglmm[1:2] %>% 
  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
auc_t780_adglmm <- lapply(auc_t780_adglmm, function(x){x@y.values[[1]]}) %>% unlist()

# up to 1020
auc_t1020_adglmm <- adglmm_pred_t1020$newdata2 %>%
  # filter(!id %in% skip_id) %>%
  mutate(sind=sind*J) %>%
  mutate(window = cut(sind, breaks = c(1020, 1260, 1440), 
                      labels = c("5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
auc_t1020_adglmm <- split(auc_t1020_adglmm, f=auc_t1020_adglmm$window)
auc_t1020_adglmm <- auc_t1020_adglmm[1] %>% 
  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
auc_t1020_adglmm <- lapply(auc_t1020_adglmm, function(x){x@y.values[[1]]}) %>% unlist()

@@

### AUC

@@

## Failed Laplace approximation

- This problem seems to be solved after changing the optimizer? (Method = "BFGS")
- The new optimizer also took less time
- The old one was SPG (large scale optimization). The new one is BFGS (quasi-newton, the same in glmmTMB)

@@

- What would be the reference method to compare our performance to? Original GLMMadaptvie perhaps is feasible on the entire dataset. But what should we use if we wanna do sub-sample comparison? Should we even do that for this data application (or simulation alone)? 
<!-- - There are still three participants with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects.  -->
<!-- - Is interpolation really necessary? It does not improve perfor- Should I calculate subject-wise AUC and average them? But some subjects are all zero within the prediction window, making AUC for this single subjects impossible to calculate...
